Investigating Biases Associated with Dietary Starch Incorporation and Retention with an Oral Biofilm Model

### 2022-05-17 analysis/paper/index.Rmd
push manuscript files to github

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  #fig.path = "frontiers/figures/",
  fig.path = "../figures/",
  fig.align = "center",
  #fig.ext = "jpg",
  fig.pos = "H",
  dpi = 300) 
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(here("analysis/paper/paper.Rmd")) 
cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor2_r <- signif(count_cor$estimate, 3)
cor2_ci <- signif(count_cor$conf.int, 3)
```

@@

Dental calculus has proven to contain a wealth of information on the
dietary habits of past populations. These insights have, to a large
extent, been obtained by the extraction and identification of starch
granules contained within the mineralised dental plaque from a wide
range of regions and time periods. The scope of previous studies have
been limited to microfossil extraction and identification to reconstruct
dietary preferences from the archaeological record, and few studies have
attempted to address the biases of starch retention in dental calculus.
Those that have considered this problem have been limited to \emph{in
vivo} studies on modern humans and non-human primates. Here, we present
a multispecies oral biofilm model, which allows experimental research on
starch incorporation and retention to be conducted on \emph{in vitro}
dental calculus in a controlled laboratory setting. The biofilms were
exposed to treatment solutions with known quantities of dietary starches
(wheat and potato) during the 25-day growth period. After this, the
starch granules were extracted from the mature biofilm (by dissolution
in EDTA), and counted. We show that the granule counts extracted from
the model dental calculus represented a low proportion (ranging from
0.06\% to 0.16\%) of the total number of granules exposed to the
biofilms throughout the experiment. Additionally, we found that the
ratios of granule sizes from the extracted starch granules differed from
the original treatment solutions, with large granules (\textgreater20
\(\mu\)m) consistently being under-represented. We also found a
correlation between the absolute granule counts and dry-weight of the
biofilm (\emph{r} = 0.659, 90\%CI{[}0.463, 0.794{]}), as well as between
the concentration (count per mg) of granules and dry-weight (\emph{r} =
0.3, 90\%CI{[}0.0618, 0.506{]}).\\
Our results reinforce previous \emph{in vivo} studies suggesting that
dental calculus presents a very small, and partly biased picture of the
original dietary intake of starches, with an over-representation of
plants producing granules smaller than 20 \(\mu\)m in size. The
experimental model presented here is well-suited to address the need for
further validation of methods and biases associated with dietary
research on dental calculus.

@@

<!-- basic introduction to the field, comprehensible to a scientist in any discipline -->
<!-- Dental calculus is mineralised plaque. And it is pretty cool. -->
Dental calculus has proven to contain a wealth of 
dietary information in the form of plant microfossils 
[@henryCalculusSyria2008; @hardyStarchGranulesDental2009], 
proteins [@hendyProteomicCalculus2018; @warinnerEvidenceMilk2014],
and other organic residues [@buckleyDentalCalculus2014]. 
This dietary information can be preserved within the mineralised dental plaque 
over many millennia, providing a unique window into the food-related behaviours of 
past populations 
[@henryCalculusSyria2008; @jovanovicNeolithicCalculus2021; @taoWheatCalculus2020] 
and extinct species [@henryNeanderthalCalculus2014; @hardyNeanderthalMedics2012].

<!-- more detailed background, comprehensible to scientists in related disciplines -->
Until recently, only a few studies directly investigated the presence of plant
microremains
in the dental calculus of archaeological remains. The ability to extract phytoliths
from the dental calculus of archaeological fauna to investigate diet was first 
noted by Armitage [-@armitageExtractionIdentification1975], 
and later by Middleton and Rovner [-@middletonOpalPhytoliths1994],
and Fox and colleagues [-@foxPhytolithCalculus1996]. Starches and phytoliths were
extracted from human dental calculus by Cummings and Magennis [-@cummingsMayanCalculus1997].  
In more recent years, the study of dental calculus has increased exponentially, 
and the wealth of information contained within the mineralised matrix has largely
been acknowledged. The use of dental calculus spans a wide
variety of archaeological research areas, such as oral microbiome 
characterisation (including pathogens) through the analysis of DNA and proteins 
[@adlerSequencingAncient2013; @warinnerPathogensHost2014],
microbotanical remains [@henryCalculusSyria2008; @hardyStarchGranulesDental2009;
@mickleburghNewInsights2012], 
other organic residues and proteins from dietary compounds 
[@buckleyDentalCalculus2014; @hendyProteomicCalculus2018], 
and nicotine use [@eerkensDentalCalculus2018]. Especially the extraction 
of starch granules has become a rich source of dietary
information, as starch granules have proven to preserve well within dental calculus
over a variety of geographical and temporal ranges 
[@pipernoStarchGrains2008; @henryNeanderthalCalculus2014; @taoWheatCalculus2020;
@jovanovicNeolithicCalculus2021].

<!-- clearly stating the general problem being addressed by this particular study -->
Despite this, our knowledge of dental calculus and the incorporation pathways of 
the various markers is limited [@radiniFoodPathways2017], as is our 
knowledge of information-loss caused by these pathways. Additionally, the methods
we use to extract and analyse dental calculus, and make inferences on past diets
represent another potential source of bias. Studies on both archaeological and
modern individuals have explored these biases in more detail.
Extraction methods were tested 
by Tromp and colleagues [-@trompEDTACalculus2017], specifically regarding
decalcification using HCl or EDTA. 
The authors found significantly more starches with the EDTA extraction method 
than the HCl extraction method; however, as noted by the authors, comparisons 
involving archaeological calculus are problematic due to variability between and 
within individuals. 
Studies conducted on modern humans [@leonardDentalCalculus2015] 
and non-human primates [@powerChimpCalculus2015], 
have explored how well microremains (phytoliths and starches)
extracted from dental calculus represent the actual dietary intake.
These studies are justifiably limited, 
despite meticulous documentation and observation, due to unknown variables and 
uncertainty involved in this kind of *in vivo* research. Dental calculus is a complex 
oral biofilm with a multifactorial aetiology and variable formation rates both 
within and between individuals [@jepsenCalculusRemoval2011; @haffajeeBiofilmPosition2009], 
contributing to
the stochasticity of starch representation being observed in numerous studies.
Additionally, the concentration of oral $\alpha$-amylase differs both between and
within individuals [@froehlichEffectOral1987; @naterHumanAmylase2005],
causing different rates of hydrolysis of the starch granules present in the oral
cavity. Add to this the effects of the many different methods
of starch processing, as well as post-depositional processes that are still being 
explored [@graneroStarchTaphonomy2020], and it becomes clear that using 
dental calculus to reconstruct diet is a highly unpredictable process.

In this exploratory study, we use an oral biofilm model to investigate the 
retention of starch granules within dental calculus in a controlled laboratory 
setting, allowing us full control over dietary input. Our main questions concern
the representation of granules extracted
from the calculus compared to the actual intake. How much of the original diet is 
incorporated into the calculus, and how much is recovered? 
Is there differential loss of information from specific dietary markers that affects 
the obtained dietary information, and how does this affect the representation of 
diet from extracted microremains?  
We find that, despite the absence of $\alpha$-amylase in
the model, a limited proportion of the starch input is actually 
retained in the calculus. We also observed a shift in the size ratios of individual 
starch granules that are incorporated into the calculus, and that the number of 
incorporated starch granules increases as the size of the calculus deposit 
increases.

@@

## Biofilm formation

In this study we employ a multispecies oral biofilm model following a modified
protocol from Sissons and colleagues [-@sissonsMultistationPlaque1991] and
Shellis [-@shellisSyntheticSalivaCultural1978]. In brief, a biofilm inoculated
with whole saliva was grown on a substrate suspended in artificial saliva, and
fed with sugar (sucrose). After several days of growth, the biofilm was exposed
to starch solutions. Mineralisation of the biofilm was aided by exposure to a
calcium phosphate solution. After 25 days of growth, the mineralised biofilm was
collected for further analysis. The setup comprises a polypropylene 
24 deepwell PCR plate (KingFisher 97003510) with a lid containing 24 pegs, which 
is autoclaved at 120&deg;C, 1 bar overpressure, for 20 mins. The individual pegs 
were the substrata on which the calculus grew. Using this system allowed for easy
transfer of the growing biofilm between saliva, feeding solutions, 
and mineral solutions.

The artificial saliva (AS) is a modified version of the basal medium mucin (BMM)
described by Sissons and colleagues [-@sissonsMultistationPlaque1991]. 
It contains 2.5 g/l partially purified mucin from porcine stomach (Type III, Sigma M1778), 
5 g/l trypticase peptone (Roth 2363.1), 10 g/l proteose peptone (Oxoid LP0085),
5 g/l yeast extract (BD 211921), 2.5 g/l KCl, 0.35 g/l NaCl, 1.8 mmol/l CaCl~2~, 
5.2 mmol/l Na~2~HPO~4~ [@sissonsMultistationPlaque1991], 6.4 mmol/l NaHCO~3~ 
[@shellisSyntheticSalivaCultural1978], 2.5 mg/l haemin. This is subsequently 
adjusted to pH 7 with NaOH pellets and stirring, autoclaved (15 min, 120&deg;C, 
1 bar overpressure), and supplemented with 5.8 $\mu$mol/l menadione, 5 mmol/l urea, 
and 1 mmol/l arginine [@sissonsMultistationPlaque1991].

Fresh whole saliva (WS) for inoculation was provided by a 31-year-old<!--needless to say I am much older now...--> male donor 
with no history of caries, who abstained from oral hygiene for 24 hours. No 
food was consumed two hours prior to donation and no antibiotics were taken up 
to six months prior to donation. 
The saliva was filtered through a sterilised (with sodium hypochlorite, 10--15% active chlorine) 
nylon cloth to remove particulates.
Substrata were inoculated with 1 ml/well of a two-fold dilution of WS in sterilised 
20% (v/v) glycerine for four hours at 36&deg;C, to allow attachment of the
salivary pellicle and plaque-forming bacteria. After initial inoculation, the 
substrata were transferred to a new plate containing 1 ml/well AS and incubated 
in a shaking incubator (Infors HT Ecotron) at 36&deg;C, 30 rpm. 
The inoculation process was repeated on days 3 and 5. 
AS was partially refreshed once per day and fully refreshed every three days, 
throughout the experiment, by transferring the substrata to a new plate containing 
stock AS. To feed the bacteria, the substrata were transferred to a new plate, containing 
5% (w/v) sucrose, for six minutes twice daily, except on inoculation days 
(days 0, 3, and 5), where the samples only received one sucrose treatment after
inoculation.

Starch treatments were initiated on day 9 to avoid starch granule counts being 
affected by $\alpha$-amylase hydrolysis from saliva inoculation days. 
An $\alpha$-amylase (EC 3.2.1.1) activity
assay was conducted to confirm that no amylase was present in the model before
starch treatments started. Starch treatments replaced sucrose treatments, occurring twice per day
for six minutes. The starch treatments involved transferring the substrata to a 
new plate containing a 0.25% (w/v) starch from potato (Roth 9441.1) solution, a 0.25% (w/v) starch from wheat (Sigma S5127) solution, and a 0.5% (w/v) mixture of equal
concentrations (w/v) wheat and potato. All starch treatments were created in dH~2~O
with 5% (w/v) sucrose. Before transferring biofilm samples to the starch treatment
plate, the plates were agitated to keep the starches in suspension in the
solutions. During treatments, the rpm was increased to 60 to facilitate contact
between starch granules and biofilms.

After 15 days, mineralisation was encouraged with a 
calcium phosphate monofluorophosphate urea (CPMU) solution containing 
20 mmol/l CaCl~2~, 12 mmol/l NaH~2~PO~4~, 5 mmol/l Na~2~PO~3~F, 500 mmol/l urea, 
and (0.04 g/l MgCl)
[@pearceConcomitantDepositionStrontium1987; @sissonsMultistationPlaque1991].
The substrata were submerged in 1 ml/well CPMU for six minutes, five times 
daily, in a two-hour cycle. During the mineralisation period, starch treatments
were reduced to once per day after the five CPMU treatments. This process was repeated 
for 10 days until the end of the experiment on day 24 
(see Figure \@ref(fig:protocol-fig) for an overview of the protocol).

All laboratory work was conducted in sterile conditions under a laminar flow hood 
to prevent starch and bacterial contamination. Control samples that only received
sucrose as a treatment were included to detect starch contamination from the
environment or cross-contamination from other wells in the same plate.

## Amylase activity detection

An $\alpha$-amylase (EC 3.2.1.1) activity assay was conducted on artificial
saliva samples collected from the plate wells on days 3, 6, 8, 9, 10, 12, and 14. 
Whole saliva samples were collected on days 0, 3, and 5 as positive controls.
Collected samples were stored at 4&deg;C until the assay was conducted on day 18. 
All samples and standard curves were run in triplicates on two separate plates.
Positive control saliva samples were compared against a standard curve containing 
H~2~O, while artificial saliva samples were compared against a standard curve
containing stock AS (due to the colour of artificial saliva).
Two photometric readings were conducted for each plate with a 540 nm filter on a 
Multiskan FC Microplate Photometer (Thermo Scientific 51119000). 
The protocol is a modified version of an Enzymatic Assay of $\alpha$-Amylase
(<https://www.sigmaaldrich.com/NL/en/technical-documents/protocol/protein-biology/enzyme-activity-assays/enzymatic-assay-of-a-amylase>) [@bernfeldAmylase1955], which measures the amount of
maltose released from starch by $\alpha$-amylase activity. Results are reported
in units (U) per mL enzyme, where 1 U releases 1 $\mu$mole of maltose in 6 minutes.
The detailed protocol can be found here: <https://www.protocols.io/view/amylase-activity-bw8jphun>.

## Treatment solutions

A 1 ml aliquot of each starch solution was taken, from which 10 $\mu$l was mounted 
on a microscope slide with an 18 x 18 mm coverslip, and counted under a light microscope
(Zeiss Axioscope A1). For wheat and mixed treatment samples, we counted three 
slide transects (at ca. 1/4, 1/2, and 3/4 of the slide), and the sample counts 
were extrapolated to the total number of granules exposed to the samples over 16
days of treatments (see Supplementary Material for more details). For potato
treatment samples, the whole slide was counted.

## Extraction method

Extraction of starches from the calculus samples was performed by dissolving the
calculus in 0.5 $\tiny{M}$ ethylenediaminetetraacetic acid (EDTA) 
[@trompEDTACalculus2017; @lemoyneCalculusPretreatments2021; @modiCalculusMethodologies2020], 
and vortexing for 3 days until the sample was completely dissolved. 
Twenty $\mu$l of sample was mounted onto a slide with an 18x18 mm coverslip.
When transferring the sample to the slide, the sample was homogenised using 
the pipette to ensure that the counted transects were representative of the 
whole slide. The count from the slide was extrapolated to the whole sample
(see Supplementary Material for more detail).

Both wheat and potato granules were divided into three size categories: 
small (<10 $\mu$m), medium (10 -- 20 $\mu$m), and large (>20 $\mu$m).

## Statistical analysis

Statistical analysis was conducted in `r version$version.string`  [@Rbase] and
the following packages: tidyverse [@tidyverse2019], broom [@R-broom], 
here [@R-here], and patchwork [@R-patchwork].
<!-- R packages used -->

To see if biofilm growth was differently affected by starch treatments, a 
one-way ANOVA with sample weight as the dependent variable (DV) and treatment 
as the grouping variable (GV) was conducted.
To analyse granule counts and calculate size proportions, mean counts for each 
treatment were taken across
both experimental plates, resulting in a mean count for each
granule size category within each treatment.

Pearson's *r* was conducted on sample weight and total starch count, as well as sample
weight and starch count per mg calculus. The total count for each sample within a
treatment was standardised by z-score to account for the differences in magnitude
between the potato and wheat counts.
This was applied to total biofilm weight and starch count per mg 
calculus (also z-score standardised) to account for differences in starch 
concentration in the calculus [as per 
@wesolowskiEvaluatingMicrofossilContent2010].

@@

All samples yielded sufficient biofilm growth and starch incorporation to be 
included in the analysis (Figure \@ref(fig:microscope-fig)), resulting in a total of 48 biofilm samples (two plates of 24),
45 of which were used for analysis (three samples were set aside for later
 analysis).
Most control samples contained no starch granules, while some contained negligible
quantities (see Supplementary Material).

## No amylase activity detected in the model

```{r amylase-activity}
# table of results reported in units amylase per mL enzyme (but let's be honest,
  # ...it doesn't really matter what the unit is. No activity is no activity)
cols <- c("1", "2", "3")  
rows <- c("S1", "S2", "S3", "B1", "B2", "B3", "B4", "B5", "BT1", "BT2", "BT3")
plt1_ph1_result <- rbind(sal1_ph1, bmm1_ph1)
  plt1_ph1_result[plt1_ph1_result < 0] <- 0

amylase_results <- rbind(sal1_ph1, sal1_ph2, sal2_ph1, sal2_ph2, bmm1_ph1,
                         bmm1_ph2, bmm2_ph1, bmm2_ph2)
amylase_results[amylase_results < 0] <- NA
positive_amylase <- na.omit(amylase_results)
```

No $\alpha$-amylase activity was detected in any of the artificial
saliva samples from any of the days that were sampled. Only positive controls
(saliva)
contained amylase activity that could be detected in the assay, ranging from
`r paste(signif(range(positive_amylase),3), collapse = " to ")` U/mL enzyme 
(full results can be found in the Supplementary Material).
The results are not comparable to other studies presenting $\alpha$-amylase activity
levels in humans, as the unit definition may differ; 
however, they are sufficient to show that there is no activity in the model.

## Treatment type had minimal effect on biofilm growth

A one-way ANOVA suggests that the type of starch used during the biofilm 
growth period had a minimal effect on the growth of the biofilm 
(expressed as total dry weight of the sample), F(`r paste(treat_df[1], treat_df[2], sep = ", ")`)
= `r treat_f.stat`, p = `r treat_p`.
A summary of sample weights is available in Table \@ref(tab:anova-tbl).

## Starch counts

<!-- starch counts in the solutions -->

```{r solution-count-out}
sol_tbl <- sol_comb_out %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(x / total, 0.1), ")")))

sol_tbl[1,3] <- NA
```

<!-- starch counts in the samples -->

```{r sample-count}
samp_counts_per <- samp_counts %>%
  mutate(across(c(s, m, l, total), 
                function(x) 
                  paste0(round(x), " (", scales::percent(signif(x / total, 3)), 
                         ")")))

# combine samp_counts_per with samp_sd_wide
samp_counts_tbl <- samp_counts_per %>%
  bind_cols(select(samp_sd_wide, c(s_sd, m_sd, l_sd, total_sd)), .name_repair = "unique") %>%
  select(treatment, starch, s, s_sd, m, m_sd, l, l_sd, total, total_sd)

samp_counts_tbl[1,3] <- NA
```

It was not possible to differentiate between potato and 
wheat starches smaller than ca. 10 $\mu$m. These were counted as wheat,
as we assumed that the majority of the small granules were wheat. We make this 
assumption based on the counts of small starches in the wheat-only and potato-only
solutions. Of the combined amount of small starches in these two solutions, 
`r paste0(signif((sol_comb_out$s[5] / sum(sol_comb_out$s[4], sol_comb_out$s[5])) * 100, 3), "%")`
are from wheat.
<!-- count of small granules from wheat solution divided by the sum of small granules 
from the potato and wheat solutions -->

<!-- absolute solution counts compared to absolute sample counts. does exposure
to higher quantities of starches mean more incorporated starches? -->
The separate wheat and potato solutions were made with a 0.25% (w/v) starch
concentration, while the mixed-starch solution was made with 0.25% (w/v) of each
starch, with a total concentration of 0.50% (w/v).
The mixed treatment had the highest absolute count of starch granules in solution (mean = `r signif(sol_comb_row$total, 3)`),
while the biofilms exposed to the wheat solution preserved the greatest number of
granules 
(mean = `r signif(subset(samp_counts, treatment == "wheat")$total, 3)`). 
The potato treatment had the lowest absolute counts in both the solution 
(`r signif(subset(sol_comb_out, treatment == "potato")$total, 3)`)
and in the biofilm samples 
(`r signif(subset(samp_counts, treatment == "potato")$total, 3)`) 
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).

### Proportion of available starches incorporated in samples

```{r sample-perc}
prop_wide <- prop %>%
  ungroup() %>%
  pivot_wider(id_cols = c(treatment, starch), 
              names_from = size, 
              values_from = prop) %>%
  add_row(prop_comb_row, .after = 2)
```

The proportion of total starches from the solutions that were incorporated 
into the samples ranged from 
`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`,
with potato granules being more readily incorporated than wheat in both the
separated- and mixed-treatment samples
(Table \@ref(tab:sample-prop-tbl)). 
There is an inverse relationship between the absolute starch count in the solutions 
and the proportional incorporation of starches in the biofilm samples, i.e., potato 
had the lowest absolute count in solutions, but the highest proportional 
incorporation, and vice versa for the mixed treatment.

<!-- mixed treatments -->
Wheat incorporation was most affected in the mixed-treatment samples, with only
`r paste0(perc_tbl$total[2])` of the total available starches being incorporated into
the sample, compared to
`r paste0(perc_tbl$total[4])`
in the separated wheat treatment.

### Size ratios differ between solutions and samples

<!-- size proportions -->
<!-- pie charts -->

<!-- overall size trends -->

```{r size-trends}
sol_size_prop <- sol_comb_out %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "solution")

samp_size_prop <- samp_counts %>%
  group_by(treatment, starch) %>%
  #select(!total) %>%
  mutate(across(where(is.numeric), function(x) x / total)) %>%
  mutate(type = "sample")

size_diff <- sol_size_prop %>%
  full_join(samp_size_prop) %>%
  pivot_longer(cols = c(s,m,l,total), names_to = "size", values_to = "count") %>%
  pivot_wider(names_from = type, values_from = count) %>%
  mutate(diff = (sample - solution) * 100) %>%
  select(!c(solution, sample)) %>%
  pivot_wider(names_from = size, values_from = diff) %>%
  select(!total)
```

Overall, medium starch granules had a higher mean rate of incorporation 
(`r scales::percent(mean(prop_wide$m, na.rm = T), 0.001)`) 
than small 
(`r scales::percent(mean(prop_wide$s, na.rm = T), 0.001)`) 
and large 
(`r scales::percent(mean(prop_wide$l, na.rm = T), 0.001)`) 
starch granules across all treatments, while large potato starches had the lowest 
rate of incorporation across all treatments.

<!-- separated and mixed treatments -->
The difference in incorporation between the size categories resulted in a change
in size ratios between the original starch solutions and the extracted samples.
Large potato granules (> 20 $\mu$m) were most affected, with a 
`r paste0(abs(signif(size_diff$l[4], 3)), "%")` 
decrease in relative abundance in the potato-only treatment, and a
`r paste0(abs(signif(size_diff$l[1], 3)), "%")` 
decrease in mixed treatments. Medium granules increased in relative abundance 
across all samples, while small granules decreased in wheat treatments and 
increased in potato treatments 
(Figure \@ref(fig:ratio-plots)).

<!-- Proportion and size ratios are the most important points to be made -->

### Biofilm weight correlated positively with extracted starch counts

```{r cor, echo=FALSE}
# prepare cor output for reporting
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor_p <- if(count_cor$p.value > 0.001){
  paste("=", signif(count_cor$p.value, 3))
} else {
  paste("< 0.001")
}

cor2_r <- signif(starch_cor$estimate, 3)
cor2_ci <- signif(starch_cor$conf.int, 3)
cor2_p <- if(starch_cor$p.value > 0.001){
  paste("=", signif(starch_cor$p.value, 3))
} else {
  paste("< 0.001")
}
# Convert r correlation result to text
direct_cor <- ifelse(cor_r < 0, "negative", "positive")
strength_cor <- if(cor_r >= 0.8){
  "very strong"
} else if(cor_r < 0.8 & cor_r >= 0.6){
  "strong"
} else if(cor_r < 0.6 & cor_r >= 0.4){
  "moderate"
} else if(cor_r < 0.4 & cor_r >= 0.2){
  "weak"
} else{
  "very weak"
}
```

Pearson's *r* suggests a 
`r strength_cor` `r direct_cor` 
correlation between the total weight of the biofilms and the total starch count 
(standardised by z-score) extracted from the samples across treatments, 
*r* = `r cor_r`, 
90%CI[`r paste(cor_ci[1], cor_ci[2], sep = ", ")`],
p `r cor_p`
(Figure \@ref(fig:cor-plot)A).

The same test was applied to total biofilm weight and starch count per mg 
calculus (also standardised by z-score), resulting in a weak positive correlation, 
*r* = `r cor2_r`, 
90%CI[`r paste(cor2_ci[1], cor2_ci[2], sep = ", ")`],
p `r cor2_p`
(Figure \@ref(fig:cor-plot)B).

@@

<!-- Big summary/important statement for the science -->
Here, we have provided a method for exploring the incorporation of dietary 
starches into the mineral matrix of a dental calculus biofilm model. Our results show
that a very low proportion of the starches exposed to the biofilm during growth are
retained in the mineral matrix, and that the size of the starch granules
may affect the likelihood of incorporation. The proportions of starch granules 
(of all sizes) present in the extracted samples were similar across all treatments
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`), 
despite large differences in absolute granule counts between wheat 
(mean = `r format(sol_comb_out$total[5], big.mark = ",")`) 
and potato
(mean = `r format(sol_comb_out$total[4], big.mark = ",")`)
solutions.  
The absolute counts, however, differed more visibly between treatments and was
proportional with the total count of granules in the treatment solutions. Wheat 
and mixed solutions had the highest absolute mean count of starch granules, and 
also had the highest absolute mean count of starch granules extracted from the 
dental calculus
(Tables \@ref(tab:solution-count-tbl) and \@ref(tab:sample-count-tbl)).
This suggests that the starches that are more frequently consumed will be present
in higher quantities in the dental calculus, at least prior to inhumation and 
degradation in the burial environment. 
Despite the low proportion of granules recovered from the model calculus 
(`r paste(min(perc_tbl$total), max(perc_tbl$total), sep = " to ")`),
the absolute counts were still substantially greater than counts recovered from 
archaeological remains 
[@trompDietaryNondietary2015; @trompEDTACalculus2017; @wesolowskiEvaluatingMicrofossilContent2010], which could in part be due to the 
lack of oral amylase activity in our model. 
<!-- Scientific findings as extension of previous work -->
Previous research conducted on dental calculus from contemporary humans and non-human
primates suggest a high level of stochasticity involved in the retention of 
starch granules in dental calculus, and that starch granules extracted from dental 
calculus are underrepresented with regard
to actual starch intake, which is consistent with our findings (illustrated by high
standard deviations and low proportional incorporation).
Leonard and colleagues [-@leonardDentalCalculus2015] found individual
calculus samples to be a poor predictor of diet in a population, as many of the 
consumed plants were missing from some individual samples, but were present in others.   
Power and colleagues [-@powerChimpCalculus2015] 
presented similar findings in non-human primates, where phytoliths were more
representative of individual diets than starch granules.
The size bias is also consistent with the findings by Power and colleagues
[-@powerChimpCalculus2015],
who found that plants producing starches 10--20 $\mu$m in
size were over-represented; however, the representation of granules larger than
20 $\mu$m in their study is unclear.

We have also shown that the size of the starch granules influences the likelihood
of incorporation into the calculus. Starch granules larger than 20 $\mu$m in
maximum length were underrepresented in the calculus samples compared to the original 
starch solutions, an effect that was consistent across all three treatments. Medium
granules (10--20 $\mu$m) were often over-represented
(Table \@ref(tab:sample-prop-tbl), 
and 
Figure \@ref(fig:ratio-plots)).
Large potato granules were most affected, potentially because of the greater
size-range. They can reach up to 100 $\mu$m in maximum length, whereas wheat 
granules generally only reach up to 35 $\mu$m 
[@gismondiStarchGranulesData2019; @haslamDecompositionStarch2004; @seidemannStarchAtlas1966, 174-176].
Granule morphology may also play a role. Large wheat granules 
are lenticular and have a larger surface area compared 
to volume, whereas large potato granules are ovoid and have a larger volume
compared to surface area 
[@vandeveldeStarchMorphology2002; @janeAnthologyStarch1994; @reichertStarchBible1913b, 364-365; @seidemannStarchAtlas1966, 174-176]. 
Another potentially important factor 
from our results is the size of the calculus deposit. We found a 
`r strength_cor` `r direct_cor` correlation between biofilm size and retained 
starch granules (Figure \@ref(fig:cor-plot)A);
a result that contradicts findings from archaeological contexts
[@wesolowskiEvaluatingMicrofossilContent2010; @dudgeonDietGeography2014].
When the concentration of starch granules
per mg calculus is considered, the correlation is weaker, but still present
(Figure \@ref(fig:cor-plot)B).
While the larger deposits contain a higher absolute count, our findings also suggest
that they contain a slightly higher concentration of starches. This may also explain
the lower mean retention of starch granules in mixed treatments compared to
wheat treatments. Wheat treatment samples (mean = `r subset(anova_tbl, treatment == "wheat")$mean` mg) were on average larger than mixed treatment samples (mean = `r subset(anova_tbl, treatment == "mix")$mean` mg) (Table \@ref(tab:anova-tbl)); and while mixed treatment solutions contained the highest mean overall
granule counts, wheat treatment samples had the highest mean starch retention.
Further research is needed to determine why this differs from previous archaeological
findings.

<!-- How are starch granules trapped in dental calculus? Does anyone actually know? 
-->
The mechanism by which starch granules are incorporated into plaque and calculus 
remains largely unknown, and few studies have directly investigated potential
mechanisms. We know that a proportion of the starch granules entering 
the mouth can become trapped in the plaque/calculus, and can be recovered from
archaeological samples of considerable age 
[@henryNeanderthalCalculus2014; @buckleyDentalCalculus2014; @wuDietEarliest2021]. 
Studies have also shown that not all starch granules come from a dietary source. 
Other pathways include cross-contamination from plant interactions in soil, such 
as palm phytoliths adhering to the skin of sweet potatoes 
[@trompDietaryNondietary2015], 
or accidental ingestion not related to food consumption 
[@radiniFoodPathways2017; @radiniMedievalWomenEarly2019].  
When starch granules enter the mouth, whether through ingestion of food or accidental
intake, they immediately encounter multiple obstacles. It is likely
that the bulk of starch granules are swallowed along with the food, and are 
only briefly present in the oral cavity. Other granules that are broken off
during mastication may be retained in the dentition through attachment to 
tooth surfaces (including plaque and dental calculus) and mucous membranes 
[@kashketFoodRetention1991; @doddsCarbohydrateRetention1988].
Bacteria also have the ability to adhere to starch granules 
[@toppingResistantStarch2003],
which would allow starches to attach to bacterial communities within the biofilm.
These granules are then
susceptible to mechanical removal by the tongue, salivary clearance, and hydrolysis
by $\alpha$-amylase [@kashketFoodParticles1996].
The susceptibility of granules to hydrolysis depends on the crystallinity and size
of the starch granule, as well as the mode of processing. Smaller and pre-processed
(e.g., cooked) starch granules are more susceptible to enzymatic degradation, 
while dehydrated starches will have a reduced susceptibility
[@lingstromStarchyFood1994; @francoStarchDegradation1992; @haslamDecompositionStarch2004; 
@bjorckStarchProcessing1984; @henryCookingStarch2009].
Cracks on the surface of the dental calculus, as well as unmineralised islands 
and channels may also be able to contain starch granules 
[@powerSEMCalculus2014; @tanCalculusUltrastructure2004; @charlierSEMCalculus2010].
Starch granules that are trapped in these pockets are (at least to some extent) 
protected from aforementioned clearance mechanisms, especially once
a new layer of plaque has covered the surface of the plaque/calculus. 
The size bias against large granules (>20 $\mu$m) from both wheat and potato 
(Table \@ref(tab:sample-prop-tbl)) may give further credence to
this incorporation pathway, as the smaller starch granules have an advantage over
larger granules, and can be stored in larger quantities.
This was also suggested by Power and colleagues [-@powerSEMCalculus2014], who
observed clusters of starches within dental calculus, rather than an even 
distribution across the surface of the dental calculus.
Granules trapped 
in plaque/calculus may still be susceptible to hydrolysis, as $\alpha$-amylase has
the ability to bind to both tooth enamel and bacteria within a biofilm and retain
a portion of its hydrolytic activity
[@scannapiecoSalivaryAmylase1993; @nikitkovaStarchBiofilms2013; @tanBacterialViability2004; @tanCalculusUltrastructure2004].
After the death of an individual, starches within dental calculus are susceptible 
to further degradation by post-depositional processes, depending on burial environment 
(pH, temperature, moisture content, microorganisms)
[@henryCookingStarch2009; @graneroStarchTaphonomy2020; @haslamDecompositionStarch2004; @francoStarchDegradation1992].
Future study should explore how burial affects the recovery of starch from the
biofilm model. <!-- I may get to it eventually, but it would be great if someone else could look at this! -->

<!-- in summary: we know very little about the mechanism behind how starches 
become trapped in dental calculus, but it would be nice to find out more. -->

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  #fig.path = "frontiers/figures/",
  fig.path = "../figures/",
  fig.align = "center",
  #fig.ext = "jpg",
  fig.pos = "H",
  dpi = 300) 
library(tidyverse)
library(broom)
library(here)
library(patchwork)
# load R scripts
source(here("analysis/scripts/analysis.R"))
source(here("analysis/scripts/tables.R"))
source(here("analysis/scripts/figures.R"))
source(here("analysis/scripts/amylase-assay.R"))
## Compile refs.bib using rbbt package ##
cite_keys <- rbbt::bbt_detect_citations(list.files(pattern = ".Rmd"))
#cite_keys <- unlist(cite_keys, use.names = FALSE)
rbbt::bbt_write_bib(here("analysis/paper/refs.bib"), keys = cite_keys, overwrite = T, translator = "bibtex")
# ## end refs.bib ##
# 
# ## Word count ##
# wd_count <- wordcountaddin::word_count()
cor_r <- signif(count_cor$estimate, 3)
cor_ci <- signif(count_cor$conf.int, 3)
cor2_r <- signif(count_cor$estimate, 3)
cor2_ci <- signif(count_cor$conf.int, 3)
```

@@

```{r child="abstract.Rmd"}

```

@@

```{r child="intro.Rmd"}

```

@@

```{r child="matmet.Rmd"}

```

@@

```{r child="results.Rmd"}

```

@@

```{r child="discussion.Rmd"}

```
